LHC abort gap monitor electronics upgrade

### 2024-04-21 bibliography.tex
added abort gap monitor acquisition system thesis to bibliography
added linke to one of the bsra status presentation
rephrased introduction
edited system architecture chapture

@@

@@

        \bibitem{bsra_calib} A. Boccardi et a1., \emph{Calibration procedures and
            automated actions for the Abort Gap Monitors of LHC}, (2015, 3), LHC-B SRA-
            ES-OOOl, \url{https:edms.cern.ch/document/1337184/1}

@@

% other sources
% https://indico.cern.ch/event/359810/contributions/853246/attachments/716754/983942/BSRA_BI_TB_Feb2015.pdf

>>>>

        \bibitem{beam_dump_system} Filippini, R., Carlier, E., Ducimetière, L., Goddard,
            B. \& Uythoven, J. \emph{Reliability Analysis of the LHC Beam Dumping System}.
            (2005,6), \url{https://cds.cern.ch/record/841092}

@@

        \bibitem{ttc_distribution} Taylor, B. \emph{TTC distribution for LHC detectors}.
            {\em IEEE Trans. Nucl. Sci.}. \textbf{45} pp. 821-828 (1998),
            \url{https://cds.cern.ch/record/363087}

@@

@@

        \bibitem{abort_gap_monitor_thesis}Pacner, P. LHC Abort Gap Monitor
            Acquisition System, Master thesis, (Brno, 2022, 59)

% other sources
% https://indico.cern.ch/event/359810/contributions/853246/attachments/716754/983942/BSRA_BI_TB_Feb2015.pdf
% https://edms.cern.ch/document/2794287/1
%

### 2024-04-21 paper.tex
edited system architecture chapture

\section{Introduction}
During the flat-top phase, the LHC ring stores up to 350 MJ of beam energy,
spread across 2808 bunches with an average beam transverse size under
\SI{1}{mm}. This high energy density can melt metallic surfaces and quench
superconducting magnets, posing a risk to the machine~\cite{LHC_report}. 
To safely extract the beam from the vacuum compartments, a series of kicker
magnets belonging to the beam dump system is used. These magnets take 3 us to
ramp up their magnetic field~\cite{beam_dump_system}. This defines the size of
an abort gap (AG) in the bunch train. Due to system imperfections, particles
populate this gap. Average AG population is in order of $10^{9}$ particles. The
population in a single bunch is four orders of magnitude higher in comparison.
When the beam dump is initiated, the particles in AG do not receive full
extraction force. That causes deflection towards the vacuum chamber. When a
certain energy accumulates in the AG, the magnet can quench. Therefore,
continuous monitoring of the AG population is necessary, which is the purpose
of the AGM~\cite{particles_in_ag}.\\
This paper reviews the upgrade of the current AGM system in the LHC since 2010,
focusing on system architecture, analog electronics chain, temperature
dependence, timing instabilities, and a comparison with the upgraded system.

\section{System architecture}
System is assambled from three major parts. First part is the optical font-end
including preamplifier placed under the vacuum chamber. Second part is a
gateware implemented on FPGA platform placed outside of the radiation area. The
last part is the software reading out measured data and providing interface to
interact with the gateware. The system architecture is depicted in
Fig.~\ref{fig:bsra_chain}, where each part is enclosed in a frame with a
number.
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{bsra_chain.pdf}
    \caption{BSRA Chain}
    \label{fig:bsra_chain}
\end{figure}

\subsection{Working principle}
Abort gap monitoring is non-invasive beam measurement with high dynamic range.
It exploits synchrotron radiation light (SL) present, when trajectory of a
particle is bend due to magnetic forces. The particle loses it's energy in a
form of light tangentally radiated to its trajectory. Any change in the light
spectrum reflects the change in the beam [REF]. 
The SL spectrum at the flat-top energy (6.8TeV) is in visible range, which
is measured by a optical front-end. This consists of photomultiplier (PMT),
converting SL to electric current, filters, and pre-amplifier to increase
magnitude before sending the signal \SI{100}{m} far away data acquisition
system (DAQ). The PMT is gate, that it opens just at the moment of abort gap.
\\
When SL is converted it is amplified and send to the DAQ. The system is
synchronous with the RF timing, that it gets information about bunch position
and turn start. The signal goes to through amplifier to the integrator, where
it is decimated by factor 10. The system is gated. Its measure window is
\SI{6.4}{\micro\second} long, but the gate opens just for the time of the abort
gap. The decimate the data to \SI{100}{ns} sample period. In other words system
measures 64 values, but just 30 have valid value. Each of these values goes to
the Exponential moving averager. This type of filter has a feedback, so the
its output reflects the evolution of the charge in the abort gap.

>>>>

\section{Introduction}
During the flat-top phase, the LHC ring stores up to \SI{350}{MJ} of beam energy,
spread across 2808 bunches with an average beam transverse size under
\SI{1}{mm}. This high energy density can melt metallic surfaces and quench
superconducting magnets, posing a risk to the machine~\cite{LHC_report}. 
To safely extract the beam from the vacuum compartments, a series of kicker
magnets belonging to the beam dump system is used. These magnets take 3 us to
ramp up their magnetic field~\cite{beam_dump_system}. This defines the size of
an abort gap (AG) in the bunch train. Due to system imperfections, particles
populate this gap. Average AG population is in order of $10^{9}$ particles. The
population in a single bunch is four orders of magnitude higher in comparison.
When the beam dump is initiated, the particles in AG do not receive full
extraction force. That causes deflection towards the vacuum chamber. When a
certain energy accumulates in the AG, the magnet can quench. Therefore,
continuous monitoring of the AG population is necessary, which is the purpose
of the AGM~\cite{particles_in_ag}.\\
This paper reviews the upgrade of the current AGM system in the LHC since 2010,
focusing on system architecture, analog electronics chain, temperature
dependence, timing instabilities, and a comparison with the upgraded system.

\section{System architecture}
The system comprises three main parts. The first part is the optical front-end
(OF), which includes a preamplifier positioned beneath the vacuum chamber. The
second part is a data acquisition system (DAQ) deployed on an FPGA platform
located outside the radiation area. The third part involves software (SW) that
retrieves, interprets, and presents the measured data, facilitating interaction
with the instrument. The system architecture is illustrated in
Fig.~\ref{fig:bsra_chain}, with each part labeled accordingly.
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{bsra_chain.pdf}
    \caption{BSRA Chain}
    \label{fig:bsra_chain}
\end{figure} 
The diagram on the left depicts the layout of the LHC, with beam 1 represented
in red and beam 2 in blue. It includes all four experiments (Alice, Atlas,
LHCb, and CMS). The AGM is situated at Point 4 (indicated by the yellow
square), close to the CMS experiment. The Timing, Trigger, and Control system
(TTC) is shown as a purple rectangle in the center of the LHC ring. This system
receives RF clocking and broadcasts it as messages on the network (TTC
distribution)~\cite{ttc_distribution}. The AGM extracts bunch clock information
(corresponding to the position of bunches) and turn clock information (marking
the start of a turn) from these messages.\\
Beam intensity is gathered by measuring the synchrotron light (SL), as
indicated by the red arrow directed towards the Multi-channel photomultiplier
(MCP-PMT). The light reaches the MCP-PMT in the optical front-end (OF).
Subsequently, the signal is transmitted over a \SI{100}{m} cable. To compensate
for losses in the cable, the signal is amplified beforehand. It then proceeds
to the Data Acquisition (DAQ) system, which measures \SI{6.4}{\micro\second}
window and integrates the intervals of \SI{100}{\nano\second}. This creates set
of 64 values published to the SW every \SI{10}{\micro\second}.

\subsection{Working principle}
The SL spectrum at the flat-top energy \SI{6.8}{TeV} is in visible band, which
is measured by a optical front-end. This consists of MCP-PMT,
converting SL to electric current, filters, and pre-amplifier to increase
magnitude before sending the signal \SI{100}{m} far away data acquisition
system (DAQ). The PMT is gate, that it opens just at the moment of abort gap. 
\\
When SL is converted it is amplified and send to the DAQ. The system is
synchronous with the RF timing, that it gets information about bunch position
and turn start. The signal goes to through amplifier to the integrator, where
it is decimated by factor 10. The system is gated. Its measure window is
\SI{6.4}{\micro\second} long, but the gate opens just for the time of the abort
gap. The decimate the data to \SI{100}{ns} sample period. In other words system
measures 64 values, but just 30 have valid value. Each of these values goes to
the Exponential moving averager. This type of filter has a feedback, so the
its output reflects the evolution of the charge in the abort gap.
